---
title: "Antibiotics"
author: "Aidan Schneider"
date: 03-09-2025
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->
# Grading Rubric
<!-- ------------------------------------------------>

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- --------------------------->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics 
df_antibiotics %>% knitr::kable()
```

# Visualization
<!-- ---------------------------------------------------->

### __q1__ Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
df_antibiotics %>%
  ggplot(aes(x = penicillin, y = streptomycin, size = neomycin, label = bacteria, color = gram)) +
  geom_point() +
  scale_color_manual(values = c("#FC4E07", "#52854C")) +
  geom_text_repel(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  ggtitle("Comparing MIC of different drugs against different bacteria")
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}

df_q12 <- df_antibiotics %>% 
  pivot_longer(cols = c(penicillin, streptomycin, neomycin),
               names_to = "antibiotic",
               values_to = "mic") %>% 
  mutate(bacteria = fct_reorder(bacteria, mic))


df_q12 %>% 
  ggplot(aes(fill = antibiotic, y= mic, x=bacteria, alpha = gram)) +
  scale_alpha_manual(values = c("positive" = 1, "negative" = 0.7)) +
  scale_y_log10() +
  geom_bar(position="dodge", stat="identity", width = 0.8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Comparing MIC values of different antibiotics against different bacteria")
  

```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}

df_13 <- df_antibiotics %>% 
  mutate(bacteria = fct_reorder(bacteria, penicillin))
  

df_13 %>% 
  ggplot(aes(x= bacteria, y = penicillin)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_log10() + 
  geom_hline(yintercept=0.1, linetype="dashed", color = "red") +
  annotate("text", x=14.8, y=0.2, label="MIC Max Acceptable Value", size=3, color="black") +
  ggtitle("Assessing the MIC values of penicilin against various bacteria")
  
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
df_long <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin, streptomycin, neomycin),
               names_to = "antibiotic",
               values_to = "mic")

df_long %>% 
  ggplot(aes(x = antibiotic, y = mic, color = antibiotic)) + 
  geom_boxplot() +
  scale_y_log10() +
  theme(legend.position = "none") + 
  labs(x = "Antibiotic", y = "MIC Values") +
  ggtitle("Assessing the range of MIC Values of Different Antibiotics")
  
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}

df_q15 <- df_antibiotics %>% 
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin), # Columns to pivot
    names_to = "antibiotic",
    values_to = "mic" 
  )

df_q15 %>% 
  ggplot(aes(x = antibiotic, y = bacteria)) +
  geom_tile(aes(fill = mic), color = "white") +
  scale_fill_gradient(
    trans = 'log10',                             
    low = "blue", high = "red"                   
  ) +
  labs(
    title = "Heatmap of MIC Values Across Antibiotics",
    x = "Antibiotic",
    y = "Bacteria",
    fill = "MIC Value"
  )

```

### __q2__ Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations*
- What is your response to the question above?
  - The most effective antibiotic appears to be neomycin
  - Streptomycin seems to be the next most effective antibiotic, while penicilin seems to be the least effective of the three
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - The most effective visual to answer this question was Visual 2. 
- Why?
  - I was able to analyze the effectiveness of each of the antibiotics by their position along the x and y axis for streptomycin and penicilin
  - for the third antibiotic i am able to determine the effectiveness based on the area of each of the points

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observations*
- What is your response to the question above?
  - Diplococcus pneumoniae was renamed Streptococcus pneumoniae as it was found to be more associated with the streptococcus genus
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - Visual 5 was most effective in helping me answer this question
- Why?
  - The Diplococcus pneumoniae seems to have more similar traits to other streptococcus bacteria, as exhibited in visuals 2, 3, and 5
  - Visual 2 shows that Diploccocus has very similar MIC values for the three types of bacteria as other streptococcus bacteria
  - Visual 3 shows a similar relationship, this time just for penicilin
  - Visual 5 shows that Diploccocus pneumoniae exhibits similar MIC values as other streptococus bacteria in the heatmap

# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
